---
title: "Getting Started with circuitscaper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with circuitscaper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

circuitscaper requires Julia (>= 1.9) with the Circuitscape and Omniscape
packages. You can install everything from R:

```{r install}
# Install circuitscaper (from GitHub during development)
# remotes::install_github("matthewkling/circuitscaper")

library(circuitscaper)

# First time only: install Julia and required packages
cs_install_julia()
```

If you already have Julia installed, `cs_setup()` will find it automatically
or you can point to it:

```{r setup}
cs_setup(julia_home = "/path/to/julia/bin")
```

## Choosing a Mode

circuitscaper provides four Circuitscape modes and one Omniscape mode. Here's
how to choose:

**Pairwise** is the most common starting point. Use it when you have discrete
sites (habitat patches, populations) and want to know how well-connected they
are to each other. The resistance matrix is useful as a distance metric for
analyses such as isolation by resistance in population genetics.

**One-to-all / All-to-one** are variants that produce different current maps.
One-to-all emphasizes how current disperses outward from each site;
all-to-one emphasizes how current converges on each site. Use these when
you care about the spatial pattern of connectivity for individual nodes,
not just the pairwise resistances.

**Advanced** is for when you want full control over source and ground
placement rather than using focal nodes. This is useful for modeling
directional movement between a defined source area and a destination.

**Omniscape** is fundamentally different — it doesn't require focal nodes at
all. It uses a moving window to compute omnidirectional connectivity
everywhere in the landscape. Use it for wall-to-wall connectivity mapping.

## Circuitscape: Pairwise Mode

The most common Circuitscape analysis computes effective resistance and
cumulative current flow between all pairs of focal nodes. You can provide
focal nodes as coordinates (a matrix or data.frame of x/y values) or as a
raster with integer IDs.

```{r pairwise}
library(terra)
library(circuitscaper)

# Resistance surface (higher values = harder to traverse)
resistance <- rast("path/to/resistance.tif")

# Option 1: Focal nodes as coordinates (simplest)
coords <- matrix(c(-120.5, 37.2,
                    -119.8, 36.9,
                    -121.1, 37.8), ncol = 2, byrow = TRUE)
result <- cs_pairwise(resistance, coords)

# Option 2: Focal nodes as a raster (integer IDs; 0 and NA are not nodes)
locations <- rast("path/to/focal_nodes.tif")
result <- cs_pairwise(resistance, locations)

# Current flow map
plot(result$current_map)

# Pairwise resistance matrix
result$resistance_matrix
```

## Circuitscape: One-to-All Mode

In one-to-all mode, each focal node takes a turn as the source while all
others are grounded simultaneously. The result includes a per-node current
map for each focal node plus a cumulative map summing across all iterations.

```{r one-to-all}
result <- cs_one_to_all(resistance, locations)
plot(result[["cumulative_current"]])
```

## Circuitscape: All-to-One Mode

All-to-one is the reverse: all other focal nodes inject current and each
node takes a turn as the ground.

```{r all-to-one}
result <- cs_all_to_one(resistance, locations)
plot(result[["cumulative_current"]])
```

## Circuitscape: Advanced Mode

Advanced mode gives full control over source and ground placement. It returns
both a current map and a voltage map.

```{r advanced}
source_layer <- rast("path/to/sources.tif")
ground_layer <- rast("path/to/grounds.tif")

result <- cs_advanced(resistance, source_layer, ground_layer)

# Access individual layers
plot(result[["cumulative_current"]])
plot(result[["voltage"]])
```

## Omniscape: Moving-Window Connectivity

Omniscape computes omnidirectional connectivity using a moving window approach.
It does not require focal nodes — every cell can serve as a source.

```{r omniscape}
result <- os_run(resistance, radius = 100, block_size = 5)

plot(result[["normalized_current"]])
plot(result[["flow_potential"]])
```

### With Source Strength Weights

```{r omniscape-source}
source_strength <- rast("path/to/habitat_quality.tif")

result <- os_run(resistance, radius = 100,
                 source_strength = source_strength,
                 block_size = 5)

plot(result[["normalized_current"]])
```

### Parallel Processing

For large landscapes, enable Julia multithreading. Julia's thread count is
fixed at startup, so set it via `cs_setup()` before any other circuitscaper
calls:

```{r omniscape-parallel}
# Set thread count at the start of your session
cs_setup(threads = 4)

result <- os_run(resistance, radius = 100,
                 block_size = 5,
                 parallelize = TRUE)
```

## Saving Outputs

By default, intermediate files are written to a temporary directory and
cleaned up. To persist output files, use `output_dir`:

```{r output-dir}
result <- cs_pairwise(resistance, locations,
                      output_dir = "my_output_directory")
```

## Tips

- **CRS preservation**: circuitscaper captures the CRS from your input raster
  and reattaches it to outputs, even though Circuitscape's ASCII grid format
  doesn't carry CRS information.

- **Solver choice**: The default `"cg+amg"` solver works well for most cases.
  For small problems, `"cholmod"` (direct solver) may be faster.

- **Verbose output**: Set `verbose = TRUE` to see Circuitscape's solver
  progress, useful for debugging or monitoring long runs.
