---
title: "Getting Started with circuitscaper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with circuitscaper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

circuitscaper requires Julia (>= 1.9) with the Circuitscape and Omniscape
packages. You can install everything from R:

```{r install}
# Install circuitscaper (from GitHub during development)
# remotes::install_github("matthewkling/circuitscaper")

library(circuitscaper)

# First time only: install Julia and required packages
cs_install_julia()
```

If you already have Julia installed, `cs_setup()` will find it automatically
or you can point to it:

```{r setup}
cs_setup(julia_home = "/path/to/julia/bin")
```

## Circuitscape: Pairwise Mode

The most common Circuitscape analysis computes effective resistance and
cumulative current flow between all pairs of focal nodes.

```{r pairwise}
library(terra)
library(circuitscaper)

# Resistance surface
resistance <- rast("path/to/resistance.tif")

# Focal nodes (integer IDs)
locations <- rast("path/to/focal_nodes.tif")

# Run pairwise analysis
result <- cs_pairwise(resistance, locations)

# Current flow map
plot(result$current_map)

# Pairwise resistance matrix
result$resistance_matrix
```

## Circuitscape: One-to-All Mode

In one-to-all mode, each focal node takes a turn as the source while all
others serve as grounds simultaneously.

```{r one-to-all}
result <- cs_one_to_all(resistance, locations)
plot(result$current_map)
```

## Circuitscape: All-to-One Mode

All-to-one is the reverse: all focal nodes are sources and each takes a turn
as the ground.

```{r all-to-one}
result <- cs_all_to_one(resistance, locations)
plot(result$current_map)
```

## Circuitscape: Advanced Mode

Advanced mode gives full control over source and ground placement.

```{r advanced}
source_layer <- rast("path/to/sources.tif")
ground_layer <- rast("path/to/grounds.tif")

result <- cs_advanced(resistance, source_layer, ground_layer,
                      write_voltage = TRUE)

# Access individual layers
plot(result[["cumulative_current"]])
plot(result[["voltage"]])
```

## Omniscape: Moving-Window Connectivity

Omniscape computes omnidirectional connectivity using a moving window approach.

```{r omniscape}
result <- os_run(resistance, radius = 100, block_size = 5)

plot(result[["normalized_current"]])
plot(result[["flow_potential"]])
```

### With Source Strength Weights

```{r omniscape-source}
source_strength <- rast("path/to/habitat_quality.tif")

result <- os_run(resistance, radius = 100,
                 source_strength = source_strength,
                 block_size = 5)

plot(result[["normalized_current"]])
```

### Parallel Processing

For large landscapes, enable Julia multithreading. Julia's thread count is
fixed at startup, so set it via `cs_setup()` before any other circuitscaper
calls:

```{r omniscape-parallel}
# Set thread count at the start of your session
cs_setup(threads = 4)

result <- os_run(resistance, radius = 100,
                 block_size = 5,
                 parallelize = TRUE)
```

## Comparing Scenarios

Use `os_compare()` to assess changes between two Omniscape runs (e.g.,
current vs. future climate).

```{r compare}
baseline <- os_run(resistance_current, radius = 100, block_size = 5)
future <- os_run(resistance_future, radius = 100, block_size = 5)

comparison <- os_compare(baseline, future)
plot(comparison[["difference"]])
plot(comparison[["ratio"]])
```

## Saving Outputs

By default, intermediate files are written to a temporary directory and
cleaned up. To persist output files, use `output_dir`:

```{r output-dir}
result <- cs_pairwise(resistance, locations,
                      output_dir = "my_output_directory")
```

## Tips

- **CRS preservation**: circuitscaper captures the CRS from your input raster
  and reattaches it to outputs, even though Circuitscape's ASCII grid format
  doesn't carry CRS information.

- **Solver choice**: The default `"cg+amg"` solver works well for most cases.
  For small problems, `"cholmod"` (direct solver) may be faster.

- **Verbose output**: Set `verbose = TRUE` to see Circuitscape's solver
  progress, useful for debugging or monitoring long runs.
